Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-03-23 paper/zhang-cook-laa-langrene-menendez.Rmd
Some changes in visual diagnostics section.

A starting point of diagnosing an optimiser is to understand how many searches an optimiser has conducted. One may want to simply plot the index value of the search points across its natural order, but a point geometry may work well if each iteration has a similar number of points. When some iterations have considerably more points than others, using a point geometry over-emphasizes the iterations that have more search points since these iterations will occupy the mast majority of the plot space. An alternative is to summarise the search in each iteration using boxplots and each iteration will then be spaced out equally. Occasionally, one may still want to switch back to a point geometry if the number of points is small in a particular iteration and this can be achieved via the `cutoff` argument in the search plot function. Additional annotations are added to facilitate better reading of the plot and these include  
1) the number of points searched in each iteration can be added as text label at the bottom of each iteration; 2) the anchor bases to interpolate are connected and highlighted in a larger size; and 3) the colour of the last iteration is in a grey scale to indicate no better basis found in this iteration.

Figure \ref{fig:toy-search} shows an example of the search plot for CRS (left) and SA (right). Both optimisers quickly find better bases in the first few iterations and then take longer to find one in the later iterations. The anchor bases, the ones found with the highest index value in each iteration, always have an increased index value in the optimiser CRS while this is not the case for SA. This feature gives SA an advantage in this simple example to quickly find the optimum.

@@

Another interesting diagnostic is to examine how the index value improves between interpolating bases since the projection on these bases is played by the tour animation. Trace plots are created by plotting the index value against time. Figure \ref{fig:toy-interp} presents the trace plot of the same optimisers as Figure \ref{fig:toy-search} and one can observe that the trace is smooth in both cases. It may seem bizarre at the first sight that the interpolation in the left panel first passes some bases with higher index value before decreases to a lower target. This happens because on the one hand, the probabilistic acceptance in SA implies that some worse bases will be accepted by the optimiser. On the other hand, guided tour interpolate between the current and target basis to provide a smooth transition between projections. These two facts indicate that a non-monotonic interpolation can't be avoided for SA. Later, in section \ref{monotonic}, there will be a discussion on improving the non-monotonic interpolation for CRS.

@@

Apart from checking the search and progression of an optimiser, looking at where the bases are positioned in the basis space is another interest. Given the orthonormality constraint, the space of projection bases $\mathbf{A}_{p \times d}$ is a $p \times d$-dimensional sphere. A dimensionality reduction method, e.g. principal component analysis is applied to first project all the bases onto a 2D space. In a projection pursuit guided tour optimisation, there are various types of bases involved: 1) The starting basis; 2) The search bases that the optimiser evaluated to produce the anchor bases; 3) The anchor bases that have the highest index value in each iteration; 4) The interpolating bases on the interpolation path; and finally 5) the end basis. The importance of these bases differs and the most important ones are the starting, interpolating and end bases. The anchor and search bases can be turned on with argument `details = TRUE`. Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign. This happens because the projection is invariant to the sign difference of the bases and so does the index value, however, this creates difficulties for comparing the optimisers since the end bases will be symmetric with respect to the origin. A sign flipping step is conducted to flip the signs of all the bases in one routine if different optimisations finish at opposite places.

@@

Figure \ref{fig:toy-pca} shows the PCA plot of CRS and PD for a 1D projection problem. Both optimisers find the optimum but PD gets closer. With PCA plot, one can visually appreciate the nature of these two optimisers: PD first evaluates points in a small neighbourhood for a promising direction, while CRS evaluates points randomly in the search space to search for the next target. There are dashed lines annotated for CRS and it describes the interruption of the interpolation, which will be discussed in details in Section \ref{monotonic}.

@@

While the previous few examples have looked at the space of 1D bases in a unit sphere, this section visualises the space of 2D bases. Recall that the columns in a 2D basis are orthogonal to each other, the space of $p \times 2$ bases is a $p$-D torus. If take $p = 3$, one would see a classical 3D torus shape as shown by the grey points in Figure \ref{fig:toy-torus}. The two circles of the torus can be observed to be perpendicular to each other and this can be linked back to the orthogonality of the columns in the bases. Two paths optimised by CRS and PD are plotted on top of the torus and coloured in green and brown respectively to match the previous few 1D space plots. The final basis found by PD and CRS are shown in a larger shape  and printed below respectively:

@@

Both optimisers have found the third variable in the first column, but the second variable in the second column differs by a sign.One would expect to see this in the torus plot as the final bases match each other when projected onto one torus circle (due to the same sign in the first column) and be symmetric when projected onto the other (due the the sign difference in the second column). In Figure \ref{fig:toy-torus}, this can be seen most clearly in frame 5 where the two circles are rotated into a line from our view.

>>>>

A starting point of diagnosing an optimiser is to understand how many searches it has conducted, i.e. we to summarise how the index is increasing over iterations and how many tries need to be sampled at each iteration. This is achieved using the search plot function: a boxplot shows the distribution of index values for each try, where the selected basis (corresponding to the highest index value) is always shown as a point. In case of few tries at a given iteration showing the data points directly may be preferred over the boxplot, this is controlled via the `cutoff` argument. Additional annotations are added to facilitate better reading of the plot and these include 1) the number of points searched in each iteration can be added as text label at the bottom of each iteration; 2) the anchor bases to interpolate are connected and highlighted in a larger size; and 3) the colour of the last iteration is in a grey scale to indicate no better basis found in this iteration.

<!-- One may want to simply plot the index value of the search points across its natural order. While such a point geometry may work well if each iteration has a similar number of points, when some iterations have considerably more points than others, it will over-emphasizes the iterations that have more search points and these iterations will occupy the vast majority of the plot space. An alternative is to summarise the search in each iteration using boxplots, with each iteration spaced out equally. Occasionally, one may still want to switch back to a point geometry if the number of points is small in a particular iteration and this can be achieved via the `cutoff` argument in the search plot function. Additional annotations are added to facilitate better reading of the plot and these include  
1) the number of points searched in each iteration can be added as text label at the bottom of each iteration; 2) the anchor bases to interpolate are connected and highlighted in a larger size; and 3) the colour of the last iteration is in a grey scale to indicate no better basis found in this iteration. 
-->

Figure \ref{fig:toy-search} shows an example of the search plot for CRS (left) and SA (right). Both optimisers quickly find better bases in the first few iterations and then take longer to find one in the later iterations. The anchor bases, the ones found with the highest index value in each iteration, always have an increased index value in the optimiser CRS while this is not the case for SA. This feature gives CRS an advantage in this simple example to quickly find the optimum.

@@

Another interesting diagnostic is to examine how the index value improves between interpolating bases since the projection on these bases is shown in the tour animation. Trace plots are created by plotting the index value against time. Figure \ref{fig:toy-interp} presents the trace plot of the same optimisations as Figure \ref{fig:toy-search} and one can observe that the trace is smooth in both cases. It may seem bizarre at the first sight that the interpolation sometimes passes bases with higher index value before it decreases to a lower target. This happens because on the one hand, the probabilistic acceptance in SA implies that some worse bases will be accepted by the optimiser. In addition, the guided tour interpolates between the current and target basis to provide a smooth transition between projections, and sometimes a higher index value will be observed along the interpolation path. This indicates that a non-monotonic interpolation can't be avoided, even for CRS. Later, in section \ref{monotonic}, there will be a discussion on improving the non-monotonic interpolation for CRS.

@@

Apart from checking the search and progression of an optimiser, looking at where the bases are positioned in the basis space is also of interest. Given the orthonormality constraint, the space of projection bases $\mathbf{A}_{p \times d}$ is a Stiefel manifold. For one-dimensional projections this forms a $p$-dimensional sphere. A dimensionality reduction method, e.g. principal component analysis is applied to first project all the bases onto a 2D space. In a projection pursuit guided tour optimisation, there are various types of bases involved: 1) The starting basis; 2) The search bases that the optimiser evaluated to produce the anchor bases; 3) The anchor bases that have the highest index value in each iteration; 4) The interpolating bases on the interpolation path; and finally 5) the end basis. The importance of these bases differs and the most important ones are the starting, interpolating and end bases. The anchor and search bases can be turned on with argument `details = TRUE`. Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign. This happens because the projection is invariant to orientation of the basis and so is the index value. However, this creates difficulties for comparing the optimisers since the end bases will be symmetric with respect to the origin. A sign flipping step is conducted to flip the signs of all the bases in one routine if different optimisations finish at opposite places.

@@

Figure \ref{fig:toy-pca} shows the PCA plot of CRS and PD for a 1D projection problem. Both optimisers find the optimum but PD gets closer. With the PCA plot, one can visually appreciate the nature of these two optimisers: PD first evaluates points in a small neighbourhood for a promising direction, while CRS evaluates points randomly in the search space to search for the next target. There are dashed lines annotated for CRS and it describes the interruption of the interpolation, which will be discussed in details in Section \ref{monotonic}.

@@

While the previous few examples have looked at the space of 1D bases in a unit sphere, this section visualises the space of 2D bases. Recall that the columns in a 2D basis are orthogonal to each other, so the space of $p \times 2$ bases is a $p$-D torus. For $p = 3$ one would see a classical 3D torus shape as shown by the grey points in Figure \ref{fig:toy-torus}. The two circles of the torus can be observed to be perpendicular to each other and this can be linked back to the orthogonality. Two paths from CRS and PD are plotted on top of the torus and coloured in green and brown respectively to match the previous plots. The final basis found by PD and CRS are shown in a larger shape  and printed below respectively:

@@

Both optimisers have found the third variable in the first direction, and the second variable in the second direction. Note however the different orientation of the basis, following from the different sign in the second column. One would expect to see this in the torus plot as the final bases match each other when projected onto one torus circle (due to the same sign in the first column) and be symmetric when projected onto the other (due the the sign difference in the second column). In Figure \ref{fig:toy-torus}, this can be seen most clearly in frame 5 where the two circles are rotated into a line from our view.
